---
title: "DSC520_Week10_Assignment_Guruprasad_VelikaduKrishnamoorthy"
author: "Guruprasad Velikadu Krishnamoorthy"
date: "2023-02-19"
output: 
  pdf_document: 
    highlight: espresso
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Gurup/GURU/Learning/Masters/Term_2/DSC520_T302_Statistics_for_Data_Science/Week_10/data")
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 90), tidy = TRUE)
```
# Week 10 Assignment 


```{r include=TRUE,message=FALSE}
library(foreign)
library(caTools)
library(dplyr)
library(kableExtra)
```

# 1.b.i Fit a binary logistic regression model to the data set that predicts whether or not the patient survived for one year (the Risk1Y variable) after the surgery. Use the glm() function to perform the logistic regression. See Generalized Linear Models for an example. Include a summary using the summary() function in your results.
```{r}
# Loading the arff file using Foreign package
thoraric_data=read.arff("ThoraricSurgery.arff")

kbl(head(thoraric_data[,1:13]),caption = "Thoraric DataFrame", booktabs = T) %>% kable_styling(latex_options = c("striped", "hold_position"))
# Examining the structure
str(thoraric_data)
# creating a split variable with 80:20 ratio for train and test datasets
set.seed(40)
split <-  sample.split(thoraric_data,SplitRatio = 0.8)
split
train <- subset(thoraric_data,split=="TRUE")
test <- subset(thoraric_data,split=="FALSE")
# Creating Logistic Regression Model 
td_model1 <- glm(Risk1Yr~.,data=train,family=binomial())
summary(td_model1)

```
# 1.b.ii. According to the summary, which variables had the greatest effect on the survival rate?
```{r}
# Reading the p-values from our model
p_values <- coef(summary(td_model1))[,4]
p_values[p_values<=0.05]
```

##### Solution:  # Results suggest based on the p values that the variables `r names(p_values[p_values<=0.05])` are statistically significant as the p values were less than 0.05.

# 1.b.iii. To compute the accuracy of your model, use the dataset to predict the outcome variable. The percent of correct predictions is the accuracy of your model. What is the accuracy of your model?
```{r}
# Running the test through the model
result1 <- predict(td_model1,train,type="response")
head(result1)

# Validate the model by creating ConfusionMatrix
confmatrix <- table(Actual_value=train$Risk1Yr,Predicted_Value=result1> 0.5)
confmatrix

# Calculating the Accuracy
accuracy_model1 <- round(((confmatrix[[1,1]]+confmatrix[[2,2]])/sum(confmatrix)),digits=4)*100
accuracy_model1

# Creating functions to calculate the Precision and Recall
model_precision <- function(input_matrix) {
	# Selecting only True positive
    tp <- input_matrix[2, 2]
	# Selecting only false positive
    fp <- input_matrix[1, 2]
    return (tp / (tp + fp))
}

model_recall <- function(input_matrix) {
# Selecting onlytrue positive
    tp <- input_matrix[2, 2]
    # Selecting only false negative
    fn <- input_matrix[2, 1]
    return (tp / (tp + fn))
}
# Calculating Precision and Recall for the model
precision_model1 <- model_precision(confmatrix)
precision_model1
recall_model1 <- model_recall(confmatrix)
recall_model1
```
##### Solution: The accuracy of the model Thoraric model is `r accuracy_model1` %, Precision is `r precision_model1` and recall is `r recall_model1`

# 2.a.Fit a logistic regression model to the binary-classifier-data.csv dataset
```{r}
# Reading the Binary Classifier Dataset
bin_clas_df <- read.csv("binary-classifier-data.csv")

kbl(head(bin_clas_df),caption = "Binary Classifier DataFrame", booktabs = T) %>% kable_styling(latex_options = c("striped", "hold_position"))
# Changing the type of label as Factor
bin_clas_df$label <- as.factor(bin_clas_df$label)
# Creating a new split variable
split_2 <-  sample.split(bin_clas_df,SplitRatio = 0.8)
# Creating seperate dataset for training and testing the model
train_2 <- subset(bin_clas_df,split=="TRUE")
test_2 <- subset(bin_clas_df,split=="FALSE")
# Creating new model for Binary Classifier dataset
bin_clas_model <- glm(label~x+y,data=train_2,family=binomial())
summary(bin_clas_model)
# Results indicate y can be significant in predicting the label as the p value is less than 0.05
```



# 2.b. The dataset (found in binary-classifier-data.csv) contains three variables; label, x, and y. The label variable is either 0 or 1 and is the output we want to predict using the x and y variables.
# What is the accuracy of the logistic regression classifier?
```{r}
# Running the test through the model
res_bin <- predict(bin_clas_model,train_2,type="response")

# Validate the model by creating a confusion matrix
confmatrix_2 <- table(Actual_value=train_2$label,Predicted_Value=res_bin> 0.5)
confmatrix_2

# Accuracy
bin_clas_model_accuracy <- round(((confmatrix_2[[1,1]]+confmatrix_2[[2,2]])/sum(confmatrix_2)),digits=4)*100
bin_clas_model_accuracy
# # Calculating Precision and Recall for the model
precision_bin_clas_model <- model_precision(confmatrix_2)
precision_bin_clas_model
recall_bin_clas_model <- model_recall(confmatrix_2)
recall_bin_clas_model
```

##### Solution: The accuracy of the model Binary Classifier data model is `r bin_clas_model_accuracy` %, Precision is `r precision_bin_clas_model` and recall is `r recall_bin_clas_model`

